---
title: 'Discovering Funders via OpenAlex'
author: "Aaron Wenger"
date: "2025-12-03"
categories:
  - R
  - Bibliometrics
draft: true
---

<!--
- future posts:
	- comparison of OA data to Crossref and Scopus
	- exploring ROR data for understanding funders
	- exploring ROR data for understanding research orgs
- making a better openalexR
	- don't lose information, keep everything if that is possible
	- allow querying and returning only the number of hits, if possible
-->

Science education research (SER) is an interesting field which operates on the borders of several other long-established fields.
It might be supposed that SER is a specialization of education research which has many public and private funders.
But no, SER is principally funded by the National Science Foundation and not by the U.S. Department of Education.
(At least in the U.S. - maybe in other countries the situation is different.)

Recently, I was asked to help develop science education research projects with the goal of securing grant funding.
The proposed projects would be in a particular topic (Nature of Science) and I wondered if I could identify all grant funders that have funded such projects.
Rather than scouring journals or conducting systematically searching relevant papers, I tried to leverage existing bibliometric data.
[OpenAlex](https://openalex.org/) is a metadatabase that centralizes and organizes such data on papers and other products (e.g., datasets) of scientific inquiry.
In this post, I show how I used OpenAlex to find all U.S. funders of SER on this topic.

# Funders in OpenAlex
OpenAlex stores various entities (works, authors, sources, etc.) with each entity type recording different metadata.
Thus, authors have affiliations and ORCID ID values while sources have abbreviated titles and ISSN numbers.
Funders are their own entity type and are linked to works (such as journal articles) via grants.
OpenAlex uses Crossref data to establish these links and then enriches funder metadata using the Research Organization Registry (ROR) and Wikidata.^[See the documentation on [funders](https://docs.openalex.org/api-entities/funders)]

As noted by the OpenAlex developers, funding metadata is currently limited, but what is available can be used freely unlike most other services.
For instance, Similar to Crossref - and thus OpenAlex - Scopus mines acknowledgment sections of papers for funding data.
The company behind Scopus, Elsevier, also provides data on grants currently being offered and those awarded using publicly available data.^[see their (site)[https://www.elsevier.com/products/funding-institutional]]
But both of these Esevier products require institutional license agreements to use.

# Fetching via OpenAlexR
```{r}
#| echo: false
library(openalexR)
library(tidyverse)
library(here)

here()
post_path <- here("posts", "working_funders_via_oa")
```

OpenAlex has a well-documented API which can be leveraged in R using the `openalexR` package.
With `openalexR`, searches can constructed using an R script and the results returned as an R object.
The main function `oa_fetch()` is used to specify a query, request it through the API, and return the results as a tibble (the default).
Here, `oa_fetch()` is used to get a full list of U.S. funders and all relevant works.
The search field `default.search` is used to specify that "nature of science" appears in the title, abstract, or fulltext (if available).
`topics.subfield.id` specifies the subfield, which in this case is "education".

```{r}
#| eval: false

funder_us <- oa_fetch(entity = "funders", country_code = "US")
papers_nos <- oa_fetch(entity = "works", default.search = '"nature of science"', topics.subfield.id = 3304, verbose = TRUE) 
```

`openalexR` is very useful in combination with the OpenAlex API documentation.
However, as of writing, the package does not return the grant (and therefore funder) field for works.
This is certainly a bug and hopefully it will be fixed soon.

# Fetching Manually
The above limitation is minor as OpenAlex has a good web interface and is generous in allowing all search results to be downloaded in one click.
After applying filters, the same set of metadata is downloaded as `.csv` file and imported into R.
(Also, results were limited to 2015 and newer, and a more restrictive "title and abstract" field, excluding full text.)

```{r}
papers <- read_csv(here(post_path, "papers_nos_education.csv")) |> print()

```

```{r}
funders <- read_csv(here(post_path, "funders_us.csv")) |> print()
```



papers_by_funder2 <- papers |>
  separate_longer_delim(cols = grants.funder_display_name, delim = "|") |>
  filter(!is.na(grants.funder_display_name)) |>
  semi_join(funders, by = join_by(grants.funder_display_name == display_name)) |>
  select(title, publication_year, type, primary_location.source.display_name, grants.funder_display_name, cited_by_count, relevance_score) |>
  print()
```

# Funder Analysis

# Final Thoughts
