---
title: "Creating and Using Relational Databases, Part 3"
author: "Aaron Wenger"
date: "2023-06-23"
categories: 
  - relational databases
  - learning R
  - data science
  - tutorial
execute: 
  echo: false
  warning: false
format:
  html:
    df-print: paged
---

```{r}
#| label: Setup and data import

library(tidyverse)
library(here)
library(knitr)
```


blah blah

## Adding MiSTEM Region and School/District Counts

Using two-table verbs from `dplyer` (`left_join()` in the code below), district counts are easily added to the ISD table using `isd_code` as a key, matching each district to its corresponding ISD.
Likewise, school counts are added to both ISD and district tables.

```{r}
#| echo: true
#| code-fold: true

## these are objects save in part 2
# write_csv(mischool_database_sch, here(filepath, "mischool_database_sch.csv"))
# write_csv(mischool_database_dst, here(filepath, "mischool_database_dst.csv"))
# write_csv(mischool_database_isd, here(filepath, "mischool_database_isd.csv"))

# mischool_database_isd <- mischool_database_isd |> 
#   left_join(
#     mischool_database_dst |> 
#       group_by(isd_code) |> 
#       summarize(count_dst = n())
#   ) |> 
#   left_join(
#     mischool_database_sch |> 
#       group_by(isd_code) |> 
#       summarize(count_sch = n())
#   )
# 
# mischool_database_dst <- mischool_database_dst |> 
#   left_join(
#     mischool_database_sch |> 
#       group_by(dst_code) |> 
#       summarize(count_sch = n())
#   )


# select and display district/school counts
```

In Michigan, the MiSTEM network is a ----------.
There are 15 MiSTEM regions ---------.
In order to add the MiSTEM region ------------.

Here's the MiSTEM region directory -------.
And adding to databases

```{r}
# region_isd <- read_csv(
#   here(filepath, "23-05-17_MiS_Region-ISD-Directory.csv"), 
#   show_col_types = FALSE
# )
# 
# region_isd
```


```{r}
# mischool_database_isd <- mischool_database_isd |> 
#   left_join(
#     region_isd |> select(included_isd_code, region_code), 
#     join_by(isd_code == included_isd_code)
#   ) |> 
#   mutate(
#     region_code = str_pad(region_code, width = 2, side = "left", pad = "0")
#   )
```

## Adding Student Count Information



```{r}
# mischool_std_count <- read_csv(
#   here(filepath, "23-05_Student-Counts_MiSchoolData.csv"), 
#   show_col_types = FALSE
# )
# 
# glimpse(mischool_std_count)
```

```{r}
# mischool_std_count_0 <- mischool_std_count |> 
#   select(
#     isd_code = ISDCode,
#     # isd_name = ISDName,
#     dst_code = DistrictCode,
#     # dst_name = DistrictName,
#     sch_code = BuildingCode,
#     sch_name = BuildingName,
#     # entity_type2_name = EntityType,
#     # county_name = CountyName,
#     stdcount_total = TOTAL_ENROLLMENT,
#     stdcount_male = MALE_ENROLLMENT,
#     stdcount_female = FEMALE_ENROLLMENT,
#     stdcount_eth_ntv = AMERICAN_INDIAN_ENROLLMENT,
#     stdcount_eth_asn = ASIAN_ENROLLMENT,
#     stdcount_eth_blk = AFRICAN_AMERICAN_ENROLLMENT,
#     stdcount_eth_ltn = HISPANIC_ENROLLMENT,
#     stdcount_eth_haw = HAWAIIAN_ENROLLMENT,
#     stdcount_eth_wht = WHITE_ENROLLMENT,
#     stdcount_eth_mxd = TWO_OR_MORE_RACES_ENROLLMENT,
#     stdcount_grd_middlecollege = EARLY_MIDDLE_COLLEGE_ENROLLMENT,
#     stdcount_grd_prk = PREKINDERGARTEN_ENROLLMENT,
#     stdcount_grd_k = KINDERGARTEN_ENROLLMENT,
#     stdcount_grd_1 = GRADE_1_ENROLLMENT,
#     stdcount_grd_2 = GRADE_2_ENROLLMENT,
#     stdcount_grd_3 = GRADE_3_ENROLLMENT,
#     stdcount_grd_4 = GRADE_4_ENROLLMENT,
#     stdcount_grd_5 = GRADE_5_ENROLLMENT,
#     stdcount_grd_6 = GRADE_6_ENROLLMENT,
#     stdcount_grd_7 = GRADE_7_ENROLLMENT,
#     stdcount_grd_8 = GRADE_8_ENROLLMENT,
#     stdcount_grd_9 = GRADE_9_ENROLLMENT,
#     stdcount_grd_10 = GRADE_10_ENROLLMENT,
#     stdcount_grd_11 = GRADE_11_ENROLLMENT,
#     stdcount_grd_12 = GRADE_12_ENROLLMENT,
#     stdcount_grd_ungraded = UNGRADED_ENROLLMENT,
#     stdcount_ecn_disadvantaged = ECONOMIC_DISADVANTAGED_ENROLLMENT,
#     stdcount_special_ed = SPECIAL_EDUCATION_ENROLLMENT,
#     stdcount_ell = ENGLISH_LANGUAGE_LEARNERS_ENROLLMENT
#   ) 
# 
# mischool_std_count_0 |> count(isd_code) |> filter(n > 1)
# mischool_std_count_0 |> count(dst_code) |> filter(n > 1)
# mischool_std_count_0 |> count(sch_code) |> filter(n > 1)
# mischool_std_count_0 |> count(isd_code, dst_code, sch_code) |> filter(n > 1)
```

```{r}
# mischool_std_count_sch <- mischool_std_count_0 |> 
#   filter(sch_name != "All Buildings") |> 
#   mutate(
#     sch_prop_ecn_disadvantaged = 
#       as.numeric(stdcount_ecn_disadvantaged) / stdcount_total,
#     sch_status_title1_15 = if_else(sch_prop_ecn_disadvantaged >= 0.15, 1, 0),
#     sch_status_title1_40 = if_else(sch_prop_ecn_disadvantaged >= 0.40, 1, 0),
#     sch_prop_eth_min = (stdcount_total - stdcount_eth_wht) / stdcount_total,
#     sch_status_majority_minority = if_else(sch_prop_eth_min >= 0.5, 1, 0)
#   )
# 
# mischool_std_count_dst <- mischool_std_count_sch |> 
#   group_by(dst_code) |> 
#   summarize(
#     stdcount_total = sum(stdcount_total),
#     stdcount_ecn_disadvantaged = sum(
#       as.numeric(stdcount_ecn_disadvantaged), na.rm = TRUE),
#     stdcount_eth_wht = sum(stdcount_eth_wht)
#   ) |> 
#   mutate(
#     dst_prop_ecn_disadvantaged = 
#       as.numeric(stdcount_ecn_disadvantaged) / stdcount_total,
#     dst_status_title1_15 = if_else(dst_prop_ecn_disadvantaged >= 0.15, 1, 0),
#     dst_status_title1_40 = if_else(dst_prop_ecn_disadvantaged >= 0.40, 1, 0),
#     dst_prop_eth_min = (stdcount_total - stdcount_eth_wht) / stdcount_total,
#     dst_status_majority_minority = if_else(dst_prop_eth_min >= 0.5, 1, 0)
#   )
# 
# mischool_std_count_isd <- mischool_std_count_sch |> 
#   group_by(isd_code) |> 
#   summarize(
#     stdcount_total = sum(stdcount_total),
#     stdcount_ecn_disadvantaged = sum(
#       as.numeric(stdcount_ecn_disadvantaged), na.rm = TRUE),
#     stdcount_eth_wht = sum(stdcount_eth_wht)
#   ) |> 
#   mutate(
#     isd_prop_ecn_disadvantaged = 
#       as.numeric(stdcount_ecn_disadvantaged) / stdcount_total,
#     isd_status_title1_15 = if_else(isd_prop_ecn_disadvantaged >= 0.15, 1, 0),
#     isd_status_title1_40 = if_else(isd_prop_ecn_disadvantaged >= 0.40, 1, 0),
#     isd_prop_eth_min = (stdcount_total - stdcount_eth_wht) / stdcount_total,
#     isd_status_majority_minority = if_else(isd_prop_eth_min >= 0.5, 1, 0)
#   )
# 
# # now adding the info above to the database tables
# mischool_database_sch_1 <- mischool_database_sch |> 
#   left_join(
#     mischool_std_count_sch |> 
#       select(
#         isd_code:sch_code, 
#         stdcount_total, 
#         sch_prop_ecn_disadvantaged:sch_status_majority_minority
#       )
#   )
# 
# mischool_database_dst_1 <- mischool_database_dst |> 
#   left_join(mischool_std_count_dst)
# 
# mischool_database_isd_1 <- mischool_database_isd |> 
#   left_join(mischool_std_count_isd)
```


## Creating Visualizations

```{r}
library(sf)
library(tidyverse)
library(here)
```

```{r}
# sf_dst <- read_sf(dsn = here("data", "School_Districts_(v17a)", "School_Districts_(v17a).shp"))
# 
# dat_dst_n <- dat_entity_1 |> 
#   group_by(dst_code) |> 
#   summarize(dst_n = n())
# 
# sf_dst1 <- sf_dst |> left_join(dat_dst_n, by = c("DCODE" = "dst_code"))
# 
# ggplot(sf_dst1, aes(fill = dst_n)) +
#   geom_sf()

```
